\begin{abstract}
Spatial epidemic models for seasonal-influenza forecasting routinely
encode inter-regional connectivity, and standard time-series methods
suggest that forecasts are improved by using one state's lagged
incidence as a covariate for another's. It is not clear, however,
whether this improvement reflects directed inter-regional coupling, or
perhaps a different phenomenon. To answer this question, we pursue two
parallel lines of analysis. First, we apply the Cram\'er-Rao Lower
Bound to a two-region SIR model and show that the connectivity
parameter is fundamentally unidentifiable under seasonal
entrainment. We validate these results on 12 seasons of excess
pneumonia-and-influenza mortality across 47 US states and observe that
CRLB-derived 95\% confidence intervals around the estimated
connectivity overlap zero for nearly all state pairs. Second, we
further validate our first analysis by investigating forecasts of
pneumonia and influenza mortality using a standard SARIMA model. We
employ two covariates: another state's pneumonia and
influenza mortality, and the national average of pneumonia and
influenza mortality. We show that the other state's influenza and
pneumonia mortality provide no improvement in forecast skill beyond
the national average. These two lines of analysis converge on the same
conclusion: connectivity cannot be inferred from the data, and does
not improve forecasts beyond the shared national-level signal. Hence,
connectivity should be dropped from forecast models used for influenza
preparedness in accordance with Occam's razor. 
\end{abstract}
