\section*{Introduction}

Inferring coupling strength between synchronized dynamical systems is a
fundamental challenge across the sciences. When oscillators are driven
by common forcing or coupled strongly enough to synchronize, their
dynamics become nearly indistinguishable---and the very coupling one
seeks to measure becomes invisible in the observations. This problem
arises in neural systems exhibiting coherent oscillations, ecological
communities driven by shared environmental cycles, and spatially
connected epidemics synchronized by seasonal forcing. Here we address
this challenge in the context of infectious disease epidemiology, where
abundant surveillance data and well-characterized dynamics provide an
ideal setting to develop and validate a theoretical framework for
assessing identifiability.

Inferring interactions from epidemic surveillance data is a central
challenge in infectious disease epidemiology. While the behavior of a
single infectious disease in an isolated population is fairly well
understood~\citep{keeling2008}, the behavior of multiple co-circulating
pathogens depends critically on their interactions~\citep{rohani2003,
  shrestha2013, domenech2019, kramer2024}. Childhood infections
exemplify the importance of such interactions: increased susceptibility
to pertussis following measles infection~\citep{coleman2015},
co-occurrence of scarlet fever and varicella~\citep{coleman2016},
immune suppression following measles causing increased mortality from
various infections~\citep{mina2015}, and competition for susceptible
individuals between measles and whooping cough~\citep{rohani2003}.
Similar complexity arises in mosquito-borne diseases, where
interactions between dengue, Zika, and chikungunya remain poorly
characterized despite extensive modeling efforts~\citep{reiner2013,
  estofolete2023}. Beyond pathogen-pathogen interactions, spatial
interactions governing disease spread between geographical regions are
equally fundamental to understanding disease ecology and designing
effective interventions.

Yet inferring interactions from surveillance data has proven remarkably
difficult. Convergent Cross Mapping (CCM), a method designed for causal
inference in nonlinear systems~\citep{sugihara2012}, was applied to
epidemiological settings to identify environmental drivers of
influenza~\citep{deyle2016} and climatic modulation of
dengue~\citep{rypdal2019}.
However,~\href{https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0169050}{Cobey
  \& Baskerville (2016)} found the method ``extremely sensitive to
periodic fluctuations, inferring interactions between independent
strains that oscillate with similar frequencies.'' In response, the developers of CCM noted that such
results may reflect ``mistaking synchrony for
causation''~\citep{sugihara2017}---a fundamental identifiability
problem that we address directly in this study. Even likelihood-based
approaches face challenges: \citet{shrestha2011} found that while some
interaction parameters were ``extremely well identified,'' others
remained poorly constrained despite extensive data. These mixed results
highlight a recurring theme: shared seasonal forcing and synchronized
dynamics can create spurious signals or mask true interactions.

A similar challenge arises in spatial epidemiology, where researchers
attempt to infer connectivity between regions from case data. If
connectivity could be reliably estimated from surveillance data alone,
it would obviate the need for external mobility data sources. Several
studies have attempted such inference, with mixed
results~\href{https://www.nature.com/articles/s41598-020-75558-9}{[Holme
    \& Rosvall, 2020]},
\href{https://link.springer.com/article/10.1007/s11538-023-01235-3}{[KhudaBukhsh
    et al., 2023]}. A recurring finding is that transmission rate and
network connectivity are fundamentally confounded: ``the product
$p\beta$ can always be estimated but the individual factors may be
unidentifiable, making it difficult to distinguish between a mildly
transmissible disease on a highly connected network from a highly
transmissible infection on a sparse
network''~\href{https://pmc.ncbi.nlm.nih.gov/articles/PMC3090691/}{[Britton
    \& O'Neill, 2011]}. More recent work confirms that ``contact patterns
and the transmission of the disease across a link are strongly related
and often are difficult to
disentangle''~\href{https://pmc.ncbi.nlm.nih.gov/articles/PMC9880946/}{[Kiss
    \& Simon, 2023]}.

These identifiability challenges have led most spatial epidemic studies
to rely on external mobility data---from commuting
surveys~\href{https://www.science.org/doi/10.1126/science.1125237}{[Viboud
    et al., 2006]}, gravity
models~\href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005382}{[Charu
    et al., 2017]}, or mobile phone
records~\href{https://www.science.org/doi/10.1126/science.abb4218}{[Kraemer
    et al., 2020]}---rather than attempting to infer connectivity from
case data. Yet such proxy measures may not accurately reflect disease
transmission pathways, and are not always
available~\href{https://pmc.ncbi.nlm.nih.gov/articles/PMC3730677/}{[Wu
    et al., 2013]}. The fundamental question remains: under what
conditions, if any, can connectivity be inferred directly from epidemic
surveillance data?

What determines whether interactions---be they between pathogens or
between regions---can be reliably inferred? Structural identifiability
analysis can determine whether parameters are theoretically recoverable
from ideal
data~\href{https://www.sciencedirect.com/science/article/abs/pii/S0025556417303164}{[Tuncer
    \& Le, 2018]},
\href{https://arxiv.org/abs/2209.12821}{[Villaverde, 2022]}. However,
structural identifiability is a binary classification that does not
quantify estimation precision. The \emph{Cram\'er-Rao Lower Bound}
(CRLB) provides exactly this: a theoretical lower bound on the variance
of any unbiased estimator, derived from the Fisher Information
Matrix~\href{https://en.wikipedia.org/wiki/Cram\%C3\%A9r\%E2\%80\%93Rao_bound}{[Cram\'er,
    1946; Rao, 1945]}. If the CRLB for a parameter is large, no
estimation method---however sophisticated---can reliably recover that
parameter from the available data. Recent work has shown that practical
unidentifiability can severely hamper epidemic
predictions~\href{https://www.science.org/doi/10.1126/sciadv.abg5234}{[Sauer
    et al., 2021]}. Computing the CRLB before optimization can save
substantial effort and prevent overinterpretation of unreliable
estimates.

Computing the CRLB requires the Jacobian matrix of model outputs with
respect to parameters, typically obtained via sensitivity equations.
For a system $\dot{x} = f(x, t, \beta)$, the sensitivities
$s = \partial x / \partial \beta$ satisfy
$\dot{s} = (\partial f / \partial x) s + \partial f / \partial \beta$,
which can be solved jointly with the original
system~\href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1009598}{[Sharpe
    et al., 2022]}. Previous implementations have used numerical
differentiation, perturbing each parameter and running multiple
simulations to approximate
derivatives~\href{https://www.sciencedirect.com/science/article/abs/pii/S0025556414001631}{[Eisenberg
    \& Hayashi, 2014]}. For continuous-time ODE models, software
packages such as
\href{https://sundials.readthedocs.io/en/latest/cvodes/Mathematics_link.html}{SUNDIALS/CVODES}
solve sensitivity equations alongside the state equations using
numerical integration. However, these approaches face practical
challenges: numerical differentiation introduces approximation errors
and requires $2p$ model evaluations for $p$ parameters; continuous-time
sensitivity analysis accumulates integration errors over long time
horizons and can suffer from numerical instabilities in stiff systems.

In this study, we take a different approach. We formulate the epidemic
model in discrete time, matching the natural resolution of weekly
surveillance data. This discrete formulation yields \emph{exact,
  closed-form update equations} for both the state variables and their
sensitivities. Our approach offers several advantages:

\begin{enumerate}
\item \textbf{Computational efficiency}: Sensitivities are computed in
  a single forward pass alongside the state, with no additional model
  evaluations required.

\item \textbf{Numerical stability}: Discrete updates avoid the
  accumulated integration errors of continuous-time solvers. There is
  no step-size selection, no stiffness to manage, and no solver
  tolerances to tune.

\item \textbf{Exact derivatives}: Unlike finite-difference
  approximations, our analytically-derived sensitivity equations
  introduce no approximation error. The Jacobian is computed to
  machine precision.

\item \textbf{Avoiding vanishing gradients}: In continuous systems,
  sensitivities can decay exponentially over long time horizons,
  particularly for stable systems. Our discrete formulation with
  bounded update coefficients (e.g., $e^{-\gamma} < 1$) maintains
  well-behaved sensitivity propagation.

\item \textbf{Natural alignment with data}: Surveillance data are
  reported in discrete intervals (daily, weekly). A discrete-time
  model matches this structure directly, avoiding the additional
  discretization step required when fitting continuous models to
  discrete observations.
\end{enumerate}

We apply this framework to a fundamental question in spatial
epidemiology: \emph{can regional connectivity be inferred from
  epidemic surveillance data?} Using a two-region SIR model with
seasonal forcing, we derive the complete sensitivity equations for the
connectivity parameter $\theta$ and compute the associated CRLB. We
then validate these theoretical predictions with multi-start
optimization experiments on synthetic data.

Our central finding is that \textbf{connectivity inference depends
  critically on epidemic synchrony}. When two regions experience
epidemics in phase---rising and falling together due to shared
seasonal forcing---the data contain almost no information about their
connectivity. Mathematically, the incidence patterns from connected
and unconnected populations become nearly indistinguishable, leading
to a large CRLB for $\theta$. In contrast, when epidemics are out of
phase (e.g., due to different seasonal timing or staggered
introductions), the differential epidemic dynamics reveal the
connectivity signal, and $\theta$ becomes identifiable.

This result provides a theoretical explanation for the difficulties
encountered in previous studies. The confounding of transmission and
connectivity reported
by~\href{https://pmc.ncbi.nlm.nih.gov/articles/PMC3090691/}{Britton \&
  O'Neill (2011)} arises precisely when epidemics are synchronized:
both regions follow similar trajectories regardless of their coupling
strength.
\href{https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0169050}{Cobey
  and Baskerville's (2016)} observation that convergent cross-mapping
fails for ``strains that oscillate with similar frequencies'' reflects
the same phenomenon. Our CRLB analysis provides a unified theoretical
foundation: synchronized dynamics destroy the Fisher Information for
coupling parameters, making inference fundamentally impossible
regardless of the statistical method employed.

Our findings have immediate practical implications. Attempts to infer
connectivity from synchronized epidemics---such as influenza across
temperate regions with similar winter peaks---are likely to fail
regardless of the estimation method employed. This explains why
studies like~\href{https://www.science.org/doi/10.1126/science.1125237}{Viboud
  et al.\ (2006)}
and~\href{https://www.science.org/doi/10.1126/science.abb4218}{Kraemer
  et al.\ (2020)} have relied on external mobility data rather than
inferring connectivity from case counts: for synchronized seasonal
epidemics, case data simply do not contain the necessary information.
Conversely, connectivity inference may be feasible when comparing
regions with different epidemic timing, such as Northern and Southern
Hemisphere locations, or during the early invasion phase of a pandemic
when regional epidemics are naturally asynchronous. More broadly, this
``look before you leap'' approach can guide study design, temper
expectations, and prevent the misinterpretation of poorly identified
parameter estimates as meaningful epidemiological quantities.

The mathematical structure underlying our results---coupled oscillators
with shared periodic forcing---arises throughout the biological and
physical sciences. Neural populations receiving common input exhibit
synchronization that may similarly obscure their connectivity. Predator-prey
systems and metapopulations driven by shared seasonality face analogous
identifiability challenges when researchers attempt to infer dispersal
or interaction strengths from population time series. Our discrete-time
CRLB framework provides a computationally efficient tool for assessing
inference feasibility in any such system. While we focus here on
epidemiological applications, where the practical stakes are high and
validation data are abundant, we anticipate that the core
insight---that synchronization fundamentally limits identifiability of
coupling parameters---will prove broadly applicable across domains where
coupled oscillatory dynamics arise.
